Drought indices of water demand and supply, soil moisture, vegetation, and its impact on LULCC in continental Chile

### 2024-01-17 article/paper_drought_Chile2023.qmd
avanzar trend

Continetal Chile has a diverse climate condition from north to south and east to west [@Aceituno2021] (@fig-studyArea), which determines its great ecosystem diversity (@fig-LCprop). The Andes Mountains are a main factor in latitudinal variation [@Garreaud2009]. To describe the climate and ecosystem of Chile, we use the Koppen-Geiger release by @Beck2023 and the landcover type persistance of 80% of times for 2001–2022, from the IGBP classification scheme [@Friedl2019] (see @sec-methods_lulc). "Norte Grande" and "Norte Chico" predominate in an arid desert climate with hot (Bwh) and cold (Bwk) temperatures. At the south of "Norte Chico," the climate changes to an arid steppe with cold temperatures (Bsk). Mainly, the land is barren, with a minor surface of vegetation types such as shrubland and grassland. In the zones "Centro" and the north half of "Sur," the main climate is Mediterranean, with warmer to hot summers (Csa and Csb). In "Centro," there is a major proportion (~50%) of chilean matorral (shrubland and savanna.[@Fuentes2020]), followed by grassland (~16%), forest (8%), and croplands (5%). The south part of "Centro" and the north part of "Austral" are dominated by an Oceanic climate (Cfb). Those zones are high in forest and grassland. The southern part of the country has a tundra climate, and in Patagonia, it is a cold semi-arid area with an extended surface of grassland.

@@

To analyze the LULCC, we use the IGBP scheme from the MCD12Q1 collection 6.1 from MODIS. This product has a yearly frequency from 2001 to 2022. The IGBP defines 17 classes; from these, we regrouped into ten macroclasses for forest, schrublands, savannas, grasslands, wetlands, croplands, urban, snow and ice, barren, and water bodies. To validate the IGBP MODIS, we compare the macroclasses with the ones of a more detailed landcover map made by @Zhao2016 for Chile for the years 2013–2014 (LCChile). The later has a spatial resolution of 30 m and three levels of defined classes; from those, we used level 1, which fits with the macroclasses derived from the IGBP MODIS. We chose the years 2013 and 2014 from MODIS IGBP because the LCChile was made with data acquired in 2013 and 2014. We resampled LCChile to the spatial resolution of the IGBP MODIS using the nearest neighbor method. Then, we took a random sample of 1000 points within continental Chile and extracted the classes that fell within each point for LCCHILE, IGP2013, and IGBP2014. Finally, we applied a confusion matrix and estimated the accuracy and F1 score as metrics of performance.

@@

The validation of IGBP2013, and IGBP2014 with LCChile reached an a ccuracy of xx11, xx2 and a F1 score of yy1,yy2 respectively (see SS1). (PONER ESTO EN LA METODOLOGIA)

@fig-zcNDVI_var and \ref{tab-landcoverTrend} show that the macrozones with major LULCC were "Sur" and "Centro" with % and % of surface evidenced change, respectively. The "Norte Chico" shows an increase in barrend land of 111 $km^2 year^{-1}$ and a reduction in the class savanna -70 $km^2 year^{-1}$. In the "Centro" and "Sur" there are changes in the chilean matorral, with an important reduction in savanna (-136 to -318 $km^2 yr^{-1}$), and an increase in shrubland and grassland. Showing a change for more dense vegetation types. It appears to be a change from the surface occupied by cropland from the "Centro" to the "Sur." Also, there is a high increase in forest (397 $Km^2 yr^{-1}$) in the "Sur," replacing the savanna lost.

@@

Regarding vegetation productivity aggregated through the macrozones, there is a negative trend in four macrozones; only "Norte Grande" shows a positive trend of 0.07 per decade (z-score). zcNDVI reached a trend of -0.07 per decade in the "Centro" and "Norte Chico," followed by "Sur" with -0.05 and "Austral" with -0.03 (@fig-zcNDVI_var). This could be associated either with a reduction in vegetation surface, a decrease in biomass, or browning [@Miranda2023]. It reached its lowest values since the year 2019, reaching an extreme condition in early 2020 and 2022 in the "Centro." If we observe the spatial variation in @fig-zcNDVI_var a, despite the fact that "Norte Grande" has a positive trend, it can be noticed that the south part has a cluster with a negative trend. The negative trend is most noticeable in "Norte Chico" and the northern part of "Centro," but from there to the south, there is an extensive surface of change in landcover class where the trend was not calculated.

```{r fig-zcNDVI_var}
#| echo: false
#| fig-pos: "!ht"
#| fig-cap: "(a) Map of the linear trend of the index zcNDVI-6 for 2001–2023. Greener colors indicate a positive trend and an increase in vegetation development; reder colors correspond to a negative trend and a decrease in vegetation development. (b) Temporal variation of zcNDVI-6 aggregated at macrozone level within continental Chile. Each horizontal panel corresponds to a macrozone from 'norte grande' to 'zona austral'."
#| out-width: "\\textwidth"
knitr::include_graphics("../output/figs/temporal_variation_zcNDVI6_macrozonas.png")
```

@@

##	Total water storage (TWS) and drought indices

>>>>

Continetal Chile has a diverse climate condition from north to south and east to west [@Aceituno2021] (@fig-studyArea), which determines its great ecosystem diversity (@fig-LCprop). The Andes Mountains are a main factor in latitudinal variation [@Garreaud2009]. To describe the climate and ecosystem of Chile, we use the Koppen-Geiger release by @Beck2023 and the landcover type persistance of 80% of times for 2001–2022, from the IGBP classification scheme [@Friedl2019] (see @sec-methods_lulc). "Norte Grande" and "Norte Chico" predominate in an arid desert climate with hot (Bwh) and cold (Bwk) temperatures. At the south of "Norte Chico," the climate changes to an arid steppe with cold temperatures (Bsk). Mainly, the land is barren, with a minor surface of vegetation types such as shrubland and grassland. In the zones "Centro" and the north half of "Sur," the main climate is Mediterranean, with warmer to hot summers (Csa and Csb). In "Centro," there is a major proportion (~50%) of chilean matorral (shrubland and savanna, [@Fuentes2021]), followed by grassland (~16%), forest (8%), and croplands (5%). The south part of "Centro" and the north part of "Austral" are dominated by an Oceanic climate (Cfb). Those zones are high in forest and grassland. The southern part of the country has a tundra climate, and in Patagonia, it is a cold semi-arid area with an extended surface of grassland.

@@

To analyze the LULCC, we use the IGBP scheme from the MCD12Q1 collection 6.1 from MODIS. This product has a yearly frequency from 2001 to 2022. The IGBP defines 17 classes; from these, we regrouped into ten macroclasses for forest, schrublands, savannas, grasslands, wetlands, croplands, urban, snow and ice, barren, and water bodies. To validate the IGBP MODIS, we compare the macroclasses with the ones of a more detailed landcover map made by @Zhao2016 for Chile for the years 2013–2014 (LCChile). The later has a spatial resolution of 30 m and three levels of defined classes; from those, we used level 1, which fits with the macroclasses derived from the IGBP MODIS. We chose the years 2013 and 2014 from MODIS IGBP because the LCChile was made with data acquired in 2013 and 2014. We resampled LCChile to the spatial resolution of the IGBP MODIS using the nearest neighbor method. Then, we took a random sample of 1000 points within continental Chile and extracted the classes that fell within each point for LCCHILE, IGP2013, and IGBP2014. Finally, we applied a confusion matrix and estimated the accuracy and F1 score as metrics of performance. The validation of IGBP2013 and IGBP2014 with LCChile reached similar metrics of performance, having an accuracy of 0.82 and a F1 score of 0.66 (see SS1).

@@

@fig-zcNDVI_var and \ref{tab-landcoverTrend} show that the macrozones with major LULCC for 2001-2022 were "Centro", "Sur", and "Austral" with 36%, 31%, and 34%, respectively. The "Norte Chico" shows an increase in barrend land of 111 $km^2 year^{-1}$ and a reduction in the class savanna -70 $km^2 year^{-1}$. In the "Centro" and "Sur" there are changes in the chilean matorral, with an important reduction in savanna (-136 to -318 $km^2 yr^{-1}$), and an increase in shrubland and grassland. Showing a change for more dense vegetation types. It appears to be a change from the surface occupied by cropland from the "Centro" to the "Sur." Also, there is a high increase in forest (397 $Km^2 yr^{-1}$) in the "Sur," replacing the savanna lost.

@@

Regarding vegetation productivity aggregated through the macrozones in the five landcover macroclasses. In "Norte Grande" there is a increase trend of 0.02 (z-index) per decade, related with types of grassland and shrubland. There is a negative trend in "Norte Chico" with -0.04  and "Centro" with -0.02 per decade. In the "Norte Chico," savanna (-0.05) has the lowest trend, and the rest of the types are around -0.04. In "Centro," shrubland reaches -0.06, followed by grassland with -0.05, and croplands and savanna have ~-0.01 per decade. This could be associated either with a reduction in vegetation surface, a decrease in biomass, or browning [@Miranda2023]. Vegetation reached its lowest values since the year 2019, reaching an extreme condition in early 2020 and 2022 in the "Centro" (Mega Drought). The "Sur" and "Austral" show a positive trend of around 0.016 per decade (@fig-zcNDVI_var). The area most affected by a negative trend in vegetation seems to be associated with the Chilean matorral [@Fuentes2021], despite the croplands being as severely impacted by drought as native vegetation does in "Norte Chico.".

```{r fig-zcNDVI_var}
#| echo: false
#| fig-pos: "!ht"
#| fig-cap: "(a) Map of the linear trend of the index zcNDVI-6 for 2001–2023. Greener colors indicate a positive trend; reder colors correspond to a negative trend and a decrease in vegetation productivity. Grey colors indicate either no vegetation or a change in landcover type for 2001–2022. (b) Temporal variation of zcNDVI-6 aggregated at macrozone level within continental Chile. Each horizontal panel corresponds to a macrozone from 'Norte Grande' to 'Austral'."
#| out-width: "\\textwidth"
knitr::include_graphics("../output/figs/temporal_variation_zcNDVI6_macrozonas_con_mapa.png")
```

@@

### 2024-01-17 article/paper_drought_Chile2023.tex
avanzar trend

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, sharp corners, interior hidden, enhanced, borderline west={3pt}{0pt}{shadecolor}, breakable, frame hidden]}{\end{tcolorbox}}\fi

@@

Continetal Chile has a diverse climate condition from north to south and
east to west \citep{Aceituno2021} (Figure~\ref{fig-studyArea}), which
determines its great ecosystem diversity (Figure~\ref{fig-LCprop}). The
Andes Mountains are a main factor in latitudinal variation
\citep{Garreaud2009}. To describe the climate and ecosystem of Chile, we
use the Koppen-Geiger release by \citep{Beck2023} and the landcover type
persistance of 80\% of times for 2001--2022, from the IGBP
classification scheme \citep{Friedl2019} (see
Section~\ref{sec-methods_lulc}). ``Norte Grande'' and ``Norte Chico''
predominate in an arid desert climate with hot (Bwh) and cold (Bwk)
temperatures. At the south of ``Norte Chico,'' the climate changes to an
arid steppe with cold temperatures (Bsk). Mainly, the land is barren,
with a minor surface of vegetation types such as shrubland and
grassland. In the zones ``Centro'' and the north half of ``Sur,'' the
main climate is Mediterranean, with warmer to hot summers (Csa and Csb).
In ``Centro,'' there is a major proportion (\textasciitilde50\%) of
chilean matorral (shrubland and savanna.\citep{Fuentes2020}), followed
by grassland (\textasciitilde16\%), forest (8\%), and croplands (5\%).
The south part of ``Centro'' and the north part of ``Austral'' are
dominated by an Oceanic climate (Cfb). Those zones are high in forest
and grassland. The southern part of the country has a tundra climate,
and in Patagonia, it is a cold semi-arid area with an extended surface
of grassland.

@@

To analyze the LULCC, we use the IGBP scheme from the MCD12Q1 collection
6.1 from MODIS. This product has a yearly frequency from 2001 to 2022.
The IGBP defines 17 classes; from these, we regrouped into ten
macroclasses for forest, schrublands, savannas, grasslands, wetlands,
croplands, urban, snow and ice, barren, and water bodies. To validate
the IGBP MODIS, we compare the macroclasses with the ones of a more
detailed landcover map made by \citep{Zhao2016} for Chile for the years
2013--2014 (LCChile). The later has a spatial resolution of 30 m and
three levels of defined classes; from those, we used level 1, which fits
with the macroclasses derived from the IGBP MODIS. We chose the years
2013 and 2014 from MODIS IGBP because the LCChile was made with data
acquired in 2013 and 2014. We resampled LCChile to the spatial
resolution of the IGBP MODIS using the nearest neighbor method. Then, we
took a random sample of 1000 points within continental Chile and
extracted the classes that fell within each point for LCCHILE, IGP2013,
and IGBP2014. Finally, we applied a confusion matrix and estimated the
accuracy and F1 score as metrics of performance.

@@

Figure~\ref{fig-zcNDVI_var} and \ref{tab-landcoverTrend} show that the
macrozones with major LULCC were ``Sur'' and ``Centro'' with \% and \%
of surface evidenced change, respectively. The ``Norte Chico'' shows an
increase in barrend land of 111 \(km^2 year^{-1}\) and a reduction in
the class savanna -70 \(km^2 year^{-1}\). In the ``Centro'' and ``Sur''
there are changes in the chilean matorral, with an important reduction
in savanna (-136 to -318 \(km^2 yr^{-1}\)), and an increase in shrubland
and grassland. Showing a change for more dense vegetation types. It
appears to be a change from the surface occupied by cropland from the
``Centro'' to the ``Sur.'' Also, there is a high increase in forest (397
\(Km^2 yr^{-1}\)) in the ``Sur,'' replacing the savanna lost.

@@

Regarding vegetation productivity aggregated through the macrozones,
there is a negative trend in four macrozones; only ``Norte Grande''
shows a positive trend of 0.07 per decade (z-score). zcNDVI reached a
trend of -0.07 per decade in the ``Centro'' and ``Norte Chico,''
followed by ``Sur'' with -0.05 and ``Austral'' with -0.03
(Figure~\ref{fig-zcNDVI_var}). This could be associated either with a
reduction in vegetation surface, a decrease in biomass, or browning
\citep{Miranda2023}. It reached its lowest values since the year 2019,
reaching an extreme condition in early 2020 and 2022 in the ``Centro.''
If we observe the spatial variation in Figure~\ref{fig-zcNDVI_var} a,
despite the fact that ``Norte Grande'' has a positive trend, it can be
noticed that the south part has a cluster with a negative trend. The
negative trend is most noticeable in ``Norte Chico'' and the northern
part of ``Centro,'' but from there to the south, there is an extensive
surface of change in landcover class where the trend was not calculated.

@@

{\centering \includegraphics{../output/figs/temporal_variation_zcNDVI6_macrozonas.png}

@@

\caption{\label{fig-zcNDVI_var}(a) Map of the linear trend of the index
zcNDVI-6 for 2001--2023. Greener colors indicate a positive trend and an
increase in vegetation development; reder colors correspond to a
negative trend and a decrease in vegetation development. (b) Temporal
variation of zcNDVI-6 aggregated at macrozone level within continental
Chile. Each horizontal panel corresponds to a macrozone from `norte
grande' to `zona austral'.}

@@

\hypertarget{total-water-storage-tws-and-drought-indices}{%
\subsection{Total water storage (TWS) and drought
indices}\label{total-water-storage-tws-and-drought-indices}}

>>>>

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, interior hidden, enhanced, frame hidden, breakable, sharp corners, boxrule=0pt]}{\end{tcolorbox}}\fi

@@

Continetal Chile has a diverse climate condition from north to south and
east to west \citep{Aceituno2021} (Figure~\ref{fig-studyArea}), which
determines its great ecosystem diversity (Figure~\ref{fig-LCprop}). The
Andes Mountains are a main factor in latitudinal variation
\citep{Garreaud2009}. To describe the climate and ecosystem of Chile, we
use the Koppen-Geiger release by \citep{Beck2023} and the landcover type
persistance of 80\% of times for 2001--2022, from the IGBP
classification scheme \citep{Friedl2019} (see
Section~\ref{sec-methods_lulc}). ``Norte Grande'' and ``Norte Chico''
predominate in an arid desert climate with hot (Bwh) and cold (Bwk)
temperatures. At the south of ``Norte Chico,'' the climate changes to an
arid steppe with cold temperatures (Bsk). Mainly, the land is barren,
with a minor surface of vegetation types such as shrubland and
grassland. In the zones ``Centro'' and the north half of ``Sur,'' the
main climate is Mediterranean, with warmer to hot summers (Csa and Csb).
In ``Centro,'' there is a major proportion (\textasciitilde50\%) of
chilean matorral (shrubland and savanna, \citep{Fuentes2021}), followed
by grassland (\textasciitilde16\%), forest (8\%), and croplands (5\%).
The south part of ``Centro'' and the north part of ``Austral'' are
dominated by an Oceanic climate (Cfb). Those zones are high in forest
and grassland. The southern part of the country has a tundra climate,
and in Patagonia, it is a cold semi-arid area with an extended surface
of grassland.

@@

To analyze the LULCC, we use the IGBP scheme from the MCD12Q1 collection
6.1 from MODIS. This product has a yearly frequency from 2001 to 2022.
The IGBP defines 17 classes; from these, we regrouped into ten
macroclasses for forest, schrublands, savannas, grasslands, wetlands,
croplands, urban, snow and ice, barren, and water bodies. To validate
the IGBP MODIS, we compare the macroclasses with the ones of a more
detailed landcover map made by \citep{Zhao2016} for Chile for the years
2013--2014 (LCChile). The later has a spatial resolution of 30 m and
three levels of defined classes; from those, we used level 1, which fits
with the macroclasses derived from the IGBP MODIS. We chose the years
2013 and 2014 from MODIS IGBP because the LCChile was made with data
acquired in 2013 and 2014. We resampled LCChile to the spatial
resolution of the IGBP MODIS using the nearest neighbor method. Then, we
took a random sample of 1000 points within continental Chile and
extracted the classes that fell within each point for LCCHILE, IGP2013,
and IGBP2014. Finally, we applied a confusion matrix and estimated the
accuracy and F1 score as metrics of performance. The validation of
IGBP2013 and IGBP2014 with LCChile reached similar metrics of
performance, having an accuracy of 0.82 and a F1 score of 0.66 (see
SS1).

@@

Figure~\ref{fig-zcNDVI_var} and \ref{tab-landcoverTrend} show that the
macrozones with major LULCC for 2001-2022 were ``Centro'', ``Sur'', and
``Austral'' with 36\%, 31\%, and 34\%, respectively. The ``Norte Chico''
shows an increase in barrend land of 111 \(km^2 year^{-1}\) and a
reduction in the class savanna -70 \(km^2 year^{-1}\). In the ``Centro''
and ``Sur'' there are changes in the chilean matorral, with an important
reduction in savanna (-136 to -318 \(km^2 yr^{-1}\)), and an increase in
shrubland and grassland. Showing a change for more dense vegetation
types. It appears to be a change from the surface occupied by cropland
from the ``Centro'' to the ``Sur.'' Also, there is a high increase in
forest (397 \(Km^2 yr^{-1}\)) in the ``Sur,'' replacing the savanna
lost.

@@

Regarding vegetation productivity aggregated through the macrozones in
the five landcover macroclasses. In ``Norte Grande'' there is a increase
trend of 0.02 (z-index) per decade, related with types of grassland and
shrubland. There is a negative trend in ``Norte Chico'' with -0.04 and
``Centro'' with -0.02 per decade. In the ``Norte Chico,'' savanna
(-0.05) has the lowest trend, and the rest of the types are around
-0.04. In ``Centro,'' shrubland reaches -0.06, followed by grassland
with -0.05, and croplands and savanna have \textasciitilde-0.01 per
decade. This could be associated either with a reduction in vegetation
surface, a decrease in biomass, or browning \citep{Miranda2023}.
Vegetation reached its lowest values since the year 2019, reaching an
extreme condition in early 2020 and 2022 in the ``Centro'' (Mega
Drought). The ``Sur'' and ``Austral'' show a positive trend of around
0.016 per decade (Figure~\ref{fig-zcNDVI_var}). The area most affected
by a negative trend in vegetation seems to be associated with the
Chilean matorral \citep{Fuentes2021}, despite the croplands being as
severely impacted by drought as native vegetation does in ``Norte
Chico.''.

@@

{\centering \includegraphics{../output/figs/temporal_variation_zcNDVI6_macrozonas_con_mapa.png}

@@

\caption{\label{fig-zcNDVI_var}(a) Map of the linear trend of the index
zcNDVI-6 for 2001--2023. Greener colors indicate a positive trend; reder
colors correspond to a negative trend and a decrease in vegetation
productivity. Grey colors indicate either no vegetation or a change in
landcover type for 2001--2022. (b) Temporal variation of zcNDVI-6
aggregated at macrozone level within continental Chile. Each horizontal
panel corresponds to a macrozone from `Norte Grande' to `Austral'.}

@@
